Matching the Time of Emergence of Transboundary Fish Stocks to Lead Time for

### 2020-11-20 Scripts/Supplements.Rmd
updates to manuscript and supplements

---
output:
  bookdown::pdf_book:
    toc : no
    keep_tex: no
    number_sections: yes
geometry: margin=1in
editor_options:
  chunk_output_type: inline
bibliography: ../References/ToE_Ref.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{float} # Use the 'float' package
- \floatplacement{figure}{H} # Make every figure with caption = h
---

<!-- --- -->
<!-- title: Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change -->
<!-- subtitle: Challenges to transboundary fisheries under climate change -->
<!-- author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, Gabriel Reygondeau^13^, U. Rashid Sumaila^1^, Colette Wabnitz^14^, and William W.L. Cheung^1^' -->
<!-- date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern.' -->
<!-- output: -->
<!--   # word_document:  -->
<!--     # reference_docx: ../References/template.docx -->
<!--   pdf_document: default -->
<!-- bibliography: ../References/ToE_Ref.bib -->
<!-- csl: ../References/science.csl -->
<!-- header-includes: -->
<!-- - \usepackage{amsmath} -->
<!-- - \usepackage{float} # Use the 'float' package -->
<!-- - \floatplacement{figure}{H} # Make every figure with caption = h -->
<!-- --- -->

<!-- # Instructions  -->

<!-- Research articles should present a major advance and can be short or long (maximum 15,000 words) format. Submissions must include an abstract (up to 150 words), an introduction, and sections with brief informative subheadings. You may include up to ten figures and/or tables and about 60 references. Supplementary materials should be limited to information that is not essential for the general understanding of the research presented in the main text and can include data sets, figures, tables, videos, or audio files.  You may include up to a total of ten figures and/or tables (combined) throughout the supplemental text.  -->

```{r setup, eval = T, echo = F, warning = F, message = F, results = 'hide'}

library(MyFunctions)

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid",
  "knitr",
 "kableExtra"
  # "here", # for dbem_import `here()`
  # "data.table", #dbem_import `fread()`
  # "readxl", # for reading excell files
  # "janitor", # for clearing names
  # "geosphere", # estimate distances between points `distm()`
  # "ggrepel",
  # "zoo", # for average mean
  # "parallel", # for mclapply,
  # "sf", # for mapping
  # "st", # for mapping
  # "rgdal", #Spatial analysis
  # "tools", #Spatial analysis 
  # "zeallot", # for Juanito's map
  # "gmt", # for estimating distances between points
  # "viridis"
)

my_lib(packages)

<!-- * corresponding author: j.palacios@oceans.ubc.ca -->

<!-- # Abstract (150  words) -->

<!-- An opening sentence that states the question/problem addressed by the research AND Enough background content to give context to the study AND A brief statement of primary results AND A short concluding sentence. -->

<!-- Climate change shifts in marine species distributions is affecting the current frameworks emplaced to manage species shared fisheries. Here, we rely on a modeling framework with multiple ensemble members to identify the time frame by which shared stocks distribution will change from historical ranges, and the change in the proportion of the distrirbution that each EEZ has under climate change. Our results suggest that 60% of shared species have emerged since 2006 in 87% of the world EEZs. By 2075 all EEZs will have an emerging species and by 2097 and all species will have emerged in at least one EEZ. Average changes in the ratio of species-share between neighboring EEZs will fluctuate between 1 and 10% with some exceeding 25% by 2030 relative to today’s ratio. Our results suggest that most changes will happen before 2030 urging policy to be re-formulated to account for the possible implications of these changes.  -->

## Introduction

Over the last century, human activities have altered the physical and biogeochemical conditions of the ocean, including warming the waters, decreasing the pH and reducing oxygen content [@IPCC:2019tn]. Distributions of marine species are closely related to the environment and the species’ preferences to environmental conditions (e.g., temperature, oxygen, salinity) [@Hutchinson:1957nt; @Nelson:2016kk]. As a result of climate change, many marine species have changed their distribution towards higher latitude, deeper water or following local environmental gradients [@Poloczanska:2016kk]. Biogeography of marine species is projected to continue to shift as ocean conditions are changing in the $21^{st}$ century under climate change [@Cheung:2010dt]. Climate change is expected to impact fisheries production, compromising our capacity to reach international sustainability goals such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Barange:2014kr; @Singh:2017ds]. The projected risks of impacts can be reduced through improving the effectiveness of current fisheries management [@Gaines:2018sg; @Cheung:2018dh], including managing fisheries on species that cross international borders, i.e., shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of  the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by States [@UnitedNations:1986tl]. As defined by the United Nation's Food and Agriculture Organization (FAO), shared stocks can be classified into three non-exclusive categories: (i) transboundary, those that cross neighboring EEZs; (ii) straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and (iii) highly migratory, mainly tunas and bill-fishes that cover vast oceanic regions including both the high seas and EEZs [@Munro:2004th]. While countries are responsible for the management of stocks within their EEZs, under UNCLOS, States are encouraged to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. A recent study estimates that, globally, there are 633 transboundary fish species, yielding an average 48.5 million tonnes of catch and USD 78 billion in fishing revenue, per year, between 2005 and 2010 [@PalaciosAbrantes:2020tv].

The effectiveness of managing fisheries targeting shared stocks is challenged by the shifts in marine species distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by]. In many cases, catch or fishing effort quota on shared stocks are based on historical records, [@Baudron:2020dj] and do not necessarily consider the biogeography of the species [@FredstonHermann:2018kp], nor the effects of climate change in the fish stocks and fisheries [@PalaciosAbrantes:2020tv]. Misalignment between fisheries resources allocation and species’ distributional shifts have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij], patterns that are expected to be exacerbated by worsening climate change [@Pinsky:2018cb]. With future shifts in species distributions under climate change, the challenges posed on shared stocks fisheries management will continue to increase in the $21^{st}$ century. Thus, there is a need to understand the time when climate change will affect the sharing distribution of these species as well as the intensity intensity of the impact in order to prepare ocean governance to species on the move [@Pinsky:2018cb; @PalaciosAbrantes:2020tv; @Link:2010ei].

The timing and intensity of climate impacts on managing transboundary fish stocks can be quantitative examined through the concepts of "Time of Emergence" (ToE) and “Thread Point” (TP) can help address some of the management challenges imposed by shifting species. Roughly, ToE is defined as the moment in time when an anthropogenic change signal (e.g., future trend) rises above the background noise of natural variability (e.g., historical variation) [@IPCC:2019tn]. This concept has been applied to investigate the emergence of multiple oceanic physical and biogeochemical variables [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja; @Schlunegger:2020gz; @Schlunegger:2019el; @Henson:2017ev]. The premise behind ToE is that we can only be confident that a significant change has been detected when the signal of anthropogenic climate change is larger than the background of natural climate variability [@Hawkins:2012hc]. The concept of thread point (TP) comes from game theory and is defined as the minimum payoff a player is willing to receive to cooperate in a game theoretic model [@Nash:1953vr]. Game theory has been widely used to investigate and manage shared stocks [@Sumaila:2013hv]. Specifically, Sumaila et al. (*in press*) applied this concept to understand the impacts of climate change to the economics of shared species between Canada and the United States.

Here, we estimate the ToE and TP of the share distribution of transboundary stocks worldwide. We rely on a species distribution model driven by output from an Earth system model with ten ensemble members to project the distribution of the 663 selected species shared [@PalaciosAbrantes:2020tv] by 280 EEZs of 198 coastal countries under a high greenhouse gas emission scenario RCP8.5 (see Methods). We treated each species in an EEZ as a single stock due to the lack of more spatially resolved data [@PalaciosAbrantes:2020tv] and only considered shared stocks between neighboring EEZs (i.e., we did not include the high seas). Our results suggest that many shared stocks have already ‘emerged’ from historical natural variability and overpassed their share TP in most neighboring EEZs posing a threat to the sustainability of shared stocks and the resilience of international treaties.

## Materials and methods

### Databases and species selection

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@Teh:2015gd; @PalaciosAbrantes:2020tv] for a total of 9,132 transboundary stocks. We defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico’s EEZ is divided into Mexico Pacific and Mexico Atlantic; Figure \@ref(fig:FigureS1)) including island territories. We determined the intersections between polygons using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by the geopolitical regions according to the United Nations and by natural biome [@Reygondeau:2019uh]. The habitat association of each species was determined based on information available from FishBase (http://www.fishbase.org). For each species and EEZ, we used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. We report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

### Projecting species distributions under climate change

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM  represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences to sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface (e.g., sea surface temperature) whereas demersal species distribution are driven by ocean bottom variables (e.g., bottom temperature). This way, the model estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid of the world (see Cheung *et al.* [-@Cheung:2010dt; -@Cheung:2016jd] for model details). The DBEM is able to project catches for the world’s large marine ecosystems and changes in catches by Exclusive Economic Zones that are consistent with observational-based estimates of catch from 1950 to 2016 [@Cheung:2016jd], Cheung et al. *in revision*].

We forced the DBEM with simulated ocean conditions from a ten ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the ‘high’ climate change emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. In GFDL-ESM2M, atmospheric surface temperature is projected to increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species’ distribution, we have to understand both the historical as well as the future climate-driven distribution variability, to infer differences between time frames. We do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, each of the ten GFDL-ESM2M-ensembles were started from infinitesimally small differences in Earth system initial conditions resulting in a unique atmosphere and ocean state at each point in time. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

### Calculating an index of transboundary range shift

We developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:FigureS2)). The centroid of a transboundary stock was determined by the average latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

<!-- $lat_{ts} = \mu (lat_{per})$ -->

<!-- $lon_{ts} = \mu (lon_{per})$ -->

\begin{equation}
\begin{split}
\label{eq:one}
lat_{ts} = \mu (lat_{per}) \\
lon_{ts} = \mu (lon_{per})
\end{split}
\end{equation}

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top $10^{th}$ percentile (e.g., per = 10%). We ran a sensitivity analysis (e.g. per = $20^{th}$, $50^{th}$, $90^{th}$ percentile) and examined the effects of different thresholds on the calculated index value (Figure \@ref(fig:FigureS2)). The centroid of a EEZ was estimated using the st package in R (version 3.2.1) (Figure \@ref(fig:FigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, we computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in R;

<!-- $$D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))$$ -->

\begin{equation}
\label{eq:two}
D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))
\end{equation}

where $lat_{eez}$ and $lat_{ts}$ are the latitudes of the EEZ and transboundary stock centroids, respectively and $lon_{eez}$ and $lon_{ts}$ are the respective longitudes.

Then, for each year we calculated the transboundary index as follows:

<!-- $$TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2$$ -->

\begin{equation}
\label{eq:three}
TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2
\end{equation}

where $D_A$ and $D_B$ represent the distance between the distribution centroids of a species and the centroid of the pair of neighbouring EEZs (A and B) sharing the species for each time step from 2006 to 2100 (*t*); and *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. Thus, a higher TI means that a shared species becomes more or less abundant relative to its neighbouring EEZs that is beyond what the fisheries experienced historically. From the perspective of a state’s fisheries, higher TI indicates an increase in the challenges to effective management and sustainability of the shared species’ fisheries. In contrast, a smaller TI suggests that the sharing of the species between neighbouring EEZs is becoming relatively more stable; a condition that favors effective management and sustainability of the shared species’ fisheries. The index was smoothed to a 10 year average to reduce interannual variability.

#### Calculating the Time-of-Emergence of the transboundary index

Knowing the point in time (e.g., year) at which the distribution of a shared stock will diverge from its natural internal variability is important in informing the lead-time to which climate adaptation in transboundary fisheries management needs to be implemented [@Link:2010ei]. Thus, we define ToE as the time when the stock shared between neighbouring EEZs will exceed the historical natural internal variability as follows:

<!-- $EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}}$ -->
<!-- $ToE = t when Emergence index[t] > 1$ -->

\begin{equation}
\begin{split}
\label{eq:four}
EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}} \\
ToE = t when Emergence index[t] > 1
\end{split}
\end{equation}

where $\mu_{TI}$ is the yearly mean TI ($\mu$) across all ten ensemble members in year *yr* and $SD_{TI}$ is the standard deviation of the smoothed (10-year moving average) TI across the ensemble members for the historical reference period (1951 - 2005). We set a threshold of TI above which we consider a shared stock to have emerged from the historical internal variability of the TI. The threshold was set as one standard deviation of the variability ($SD_{TI}$) (i.e., Equation \@ref(eq:four), representing a probability of 68% that the transboundary index has emerged from historical variability). We tested the sensitivity of the calculated emergence index to  a higher emergence threshold of two times the standard deviation (i.e., representing a probability of 95% that the index has emerged).

<!-- #### Computing a transboundary fisheries vulnerability index (VI) -->

<!-- We created a country-specific vulnerability index (VI) to examine the risk of impacts of species’ range shifts on transboundary fisheries management. The VI reflects both the rate of change of shared stocks’ distribution (indicated by the time-of-emergence of the transboundary index) and the importance of the shifting species to a country’s total catch and revenue from the species. The VI is calculated from: -->

<!-- $$ index' = (P_c + P_v)/ToE $$ -->

<!-- \begin{equation} -->
<!-- \label{eq:five} -->
<!-- VI' = \frac{(P_c + P_v)}{ToE} -->
<!-- \end{equation} -->

<!-- where $P_c$ and $P_v$ indicate the proportion accounted for by shared species in total catch and value generated from transboundary fisheries within a country’s EEZ, respectively. Here, the proportion of total catch and value are represented as the average of 2005 to 2014 within individual EEZs according to the Sea Around Us. The VI was then normalized to a 0 (less vulnerable) to 100 (more vulnerable) scale. -->

### Estimating the threat point (TP) of transboundary stock share

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., "threat point" (TP) [@Sumaila:2013hv]. Here, we define TP as the minimum proportion of a shared stock within an EEZ that a country will accept to engage in cooperative management with the country’s sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the shared stock.

First we estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. We did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ [@PalaciosAbrantes:2020tv]. Second, we average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($t_h$) spans from 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each stock. We then selected two future periods; the early $21^{st}$ century as the average of 2021 to 2040 (*te*), and the mid $21^{st}$ century as the average of 2041 to 2060 (*tm*). We chose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_h$) following Palacios-Abrantes et al. (*in press*),

<!-- $$ \Delta{SSR_{e,m}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100 $$ -->
<!-- $TP = \Delta{SSR}_{e,m} < \mu \pm \sigma$ -->

\begin{equation}
\begin{split}
\label{eq:six}
\Delta{SSR_{e,s}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100 \\
TP = \Delta{SSR}_{e,m} < \Delta{SSR}_{th} \pm \sigma
\end{split}
\end{equation}

where *e* is EEZ and *s* is stock. Finally, a substantial change in SSR of each EEZ was determined when $\Delta{SSR_{e,m}}$ was beyond one standard deviation of the historical distribution i.e., when $SSR_f \geq (SSR_f + \sigma)$ or $SSR_f \leq (SSR_f - \sigma)$. In other words, when the future SSR is below or above a neighbouring EEZ TP.

### Statistical analysis

We tested the results for normality (e.g. skewness = 1.02, kurtosis = 2.83) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary species. Specifically, we tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, our null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat associations, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software R version 3.5.2 (2018-12-20; Eggshell Igloo) with the packages data.table [@Packagedatatable:2019uh], janitor [@PackagejanitorSim:2018ur], rfishbase [@PackagerfishbaseR:2019th], sf [@PackagesfSimpleF:2018vp], sp [@PackagespClasses:2019tn], tidiverse [@PackagetidyverseE:2017vq], tidytext [@PackagetidytextTe:2019vk], gmt [@PackagegmtInterfa:2017vh] and zoo [@PackagezooSInfr:2019tm], ggrepel [@PackageggrepelAut:2020to], zeallot [@PackagezeallotMul:2018wy], viridis [@PackageviridisDef:2018us], cowplot [@CowplotStreamlined:2019wt], moments [@PackagemomensMome:2015uh] and pgirmess [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

## Results

```{r Load_Results, eval = TRUE,  echo = FALSE, include = FALSE}

prop_r <- read_csv("/Volumes/HALL2000/Data/TransEmergence/Results/proportion_change_2005.csv") %>% 
  gather("time_step","change",per_early,per_mid)

### Time of Emergence of transboundary species

The average ToE of the transboundary index (TI) across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years (Figure \@ref(fig:Figure1). Fish stocks with TI that first emerged from their historical variability in each EEZ was projected to be as early as 2006. In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. About 55% (n = 3,154) of those stocks are projected to have their shifts in distributions emerged between 2006 and 2020. Also, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century and are not expected to stop there (Figure \@ref(fig:Figure1)). The number of EEZs with emerging stocks is projected to increase steeply from 87% to 96% during the period of 2006 to 2020. After the mid-2020s, the rate of increase in the number EEZs slow down to an average of 0.74% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:Figure1)).

```{r Figure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Time of emergence by Exclusive Economic Zone and transboundary stock. A) Showing results in absolute numbers. B) Showing the result as a percentage of the total stocks and EEZ."}

img <- readPNG("../Figures/Fig1.png")
grid.raster(img)

<!-- ![Figure 1. Cumulative increase in time of emergence by Exclusive Economic Zone and transboundary species in absolute numbers and percentage.](../Figures/Fig2.png) -->

The median ToE of the transboundary index of stocks varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, X2 = 287.23, DF = 93, p < 0.001). Around one fifth (19%) of the EEZs have an average time of emergence across stocks prior to 2020 (Figure 2). Notable exceptions where the ToE is relatively late include Honduras (ToE = 2071), New Zealand (ToE = 2055), Ireland (ToE = 2067) and Pitcairn (ToE = 2070). The EEZs in Eastern Asia, Latin America, the Caribbean and Polynesia regions differed significantly (p < 0.05) from (almost) all other regions (see Table S1 for test statistics). (Figure \@ref(tab:TableS1).

Tropical latitudes also hold the majority of countries with the largest number of shifts in transboundary stocks due to climate change (Figure \@ref(fig:Figure2). Only in very few cases like Brazil and Alaska’s and Russia’s polar EEZs, no stock will emerge between 2006 and 2100. From the top 10 more vulnerable countries, seven are located in tropical waters (Tokelau 100, the	Marshall Islands 92, Peru 88, Kiribati 80, Tuvalu 80, Papua New Guinea 78, and the Bahamas 74) while only three are outside the tropics (Estonia 86, Finland 82 and Sweden 78). We set an arbitrary threshold of  1 s.d. (67% confidence) to determine the ToE (see Methods). We tested the sensitivity of our results using a larger threshold (i.e.,  2 s.d for a 95% confidence). While the selected 1 s.d. threshold is less conservative than a 2 s.d., the distribution of results do not dramatically change between thresholds, with a difference in ToE of about 4 years (Figure \@ref(fig:FigureS2)).

```{r Figure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Average time of emergence of transboundary stocks. A) Land polygon shows the vulnerability index of individual country’s fisheries management to shifts in transboundary stocks. The warmer the colour the greater the vulnerability. Exclusive Economic Zone polygon represents the average time of emergence across all stocks, with the cooler the colour the later the average time of emergence. The Density plot on the right shows the latitudinal number of emerging stocks across EEZ centroids (10 degrees latitude running mean). Aqua color = no emerging stocks. B) Time of emergence according to the United Nations sub regions. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent points outside of 1.5*IQR."}

img <- readPNG("../Figures/Fig2.png")
grid.raster(img)

<!-- ![Figure 2. Average time of emergence of transboundary specie. Land polygon shows the vulnerability index. Exclusive Economic Zone polygon represents the average time of emergence across all stocks. Density plot on the right shows the latitudinal number of emerging stocks across EEZ centroids (10^o running mean).](../Figures/Fig3.png) -->

Our analysis suggests that most (n = 47) countries responsible for 75% of the yearly revenue generated from transboundary species between 2005 and 2010 will see their distribution range shifts beyond historical variabilities before 2040 (Figure \@ref(fig:Figure3)). Particularly, nine of the top ten countries earning the most revenue from transboundary species, share a border. In some cases, like Indonesia, Thailand and the United States, stocks’ range shifts swill emerge by around 2030. The 47 countries that make up the bulk of the global fisheries revenues also have more stocks with range shifts emerging from historical variabilities.

```{r Figure3, eval = T,  echo = F, fig.width=7, fig.height=8,fig.pos= "H", fig.cap =  "Average time of emergence of transboundary species per fishing entity, aggregated by colour according to region. Fishing revenue on a logarithmic scale. Showing country names for the top 90th percentile of annual fishing revenue."}

img <- readPNG("../Figures/Fig3.png")
grid.raster(img)

<!-- ![Figure 3. Average time of emergence of transboundary species per fishing entity. Fishing revenue in logarithmic scale. Showing names for the top 75th percentile. Countries marked with an asterisk have at least one neighboring border in the group.](../Figures/Fig4.png) -->

We compared species time of emergence within EEZs by species’ habitat association (Figure \@ref(fig:Figure4)). Each species is assigned to be associated to one of the xx habitat types based on information available from FishBase (for fish, ref) or SeaLifeBase (for invertebrates, ref). We found significant differences in the time of emergence of stocks; range shifts by their habitat association (Kruskal-Wallis, $X^2$ = 286.48, DF = 93, p < 0.001). Specifically, time of emergence of range shifts of species that are associated with pelagic-oceanic and reef habitats were significantly different than other habitat preferences (See Table \@ref(tab:TableS2)). On average, species associated with pelagic oceanic habitats e,g, tunas (*Thunnus sp.*), bathy-pelagic habitats e.g., blue whiting (*Micromesistius poutassou*) and corals (e.g., Bigeye snapper *Lutjanus lutjanus*) are projected to have an earliest time of emergence (Figure \@ref(fig:Figure4)). Species associated with bathydemersal habitats such as Alaska plaice (*Pleuronectes quadrituberculatus*) generally have an average time of emergence of range shifts that is later than 2025 (Figure \@ref(fig:Figure4)).

```{r Figure4, eval = T,  echo = F, fig.width=14, fig.height=14,fig.pos= "H", fig.cap =  "Regional time of emergence per habitat association. Regions according to the United Nations and habitat association to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers outside of 1.5*IQR."}

img <- readPNG("../Figures/Fig4.png")
grid.raster(img)

<!-- ![Figure 4. Regional time of emergence per habitat association. Regions according to the United Nations and habitat association to FishBase. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers (outside of 1.5*IQR.](../Figures/Fig5.png) -->

<!-- We hypothesized that species environmental preferences (e.g. presence in certain biomes and habitats), climate change effects on water temperature (e.g. ToE of SST), human drivers (e.g. catch rate), and latitudinal gradient would have significant influence in the ToE of transboundary species (Table \ref(tab:TableS3)). Results from the multiple linear regression analysis show a significant relation between ToE of transboundary species and latitudinal gradient (*p* > 0.001), tropical (*p* = 0.013) and temperate (*p* = 0.002) biomes, EEZs in the Americas (*p* > 0.001), Asia (*p* > 0.001) and Oceania (*p* = 0.002), and some habitats (*p* > 0.001) (Table \ref(tab:Table1)). -->

```{r Table1, eval = F,  echo = F, message = F, warning  = F}

read_csv("../Tables/mlr_result.csv") %>% 
  knitr::kable(
    "latex",
    booktabs = T,
    captio= "Multiple regression results for time of emergence of transboundary species. Only showing significant variables. The model significantly predicts time of emergence (F[18,6155] = 20.68, p < 0.001, R2 = 0.05)"
  )

### Changes in the threat point of transboundary stocks

```{r analysis_results, eval = F,  echo = FALSE, include = FALSE}

# Early century
n_eez <- prop_r %>%
  filter(variable == "mean_eez" & 
           !is.na(change),
         time_step == "per_early"
         ) %>% 
  pull(eez_name) %>% 
  unique() %>% 
  length()

n_eez_per <- round(n_eez/275*100,2)

### Max values early
max_values <- prop_r %>% 
  filter(variable == "max_eez" & 
           !is.na(change),
         # time_step == "per_early"
         )

max_values$per_early

### Average
average_mid <- prop_r %>% 
  filter(variable == "mean_eez" & 
           !is.na(change)
         )
```

We estimated the changes in the SSR of transboundary stocks by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to the recent past (1951-2005) and highlighting changes below a neighbouring country’s TP. Our results suggest that 87% (n = 238) of the world EEZs will see a change in SSR below the country’s TP by 2030 (Figure \@ref(fig:Figure5)A). Globally, average changes in SSR of transboundary stocks in most EEZs are projected to be lower than 25% by the early $21^{st}$ century. The changes in SSR is expected to increase slightly towards the mid $21^{st}$ century (Figures \@ref(fig:FigureS4) and \@ref(fig:FigureS5)). However, changes in SST for specific stocks and in some EEZs are projected to be exceptionally high. For example, Guatemala is sharing *Panulirus gracilis* stock with Mexico, and Guatemala’s share is projected to increase by >50% from 12% to 26% by the early $21^{st}$ century. By the early $21^{st}$ century, the percentage of transboundary stocks per EEZ that will see a change in SSR above the country’s TP historical distribution  (e.g., $\Delta{SSR} > \mu \pm \sigma$) will be relatively large (global mean of 69% of stocks in an EEZ).

```{r Figure5, eval = T,  echo = F, fig.width=12, fig.height=10,fig.pos= "H", fig.cap =  "Changes in Specie-Share Ratio by 2030 (2021-2040) relative to 1951-2005. Lines represent the average gain of transboundary stock share ratio with arrows going from EEZ with decreasing stock share (point) to those that are gaining shares (arrowhead). Land polygons represent the percentage of initial stocks that will see a shift. Panel B zooms in to specific areas shown within the grey dashed line boundaries in A."}

img <- readPNG("../Figures/Fig5.png")
grid.raster(img)

<!-- ![Figure 1. Changes in Specie Share Ratio by 2030 (2021-2040) relative to 1951-2005. Lines represent the average gain of transboundary specie share ratio with arrows going from the losing EEZ (point) to the winner (arrowhead). Polygons represent the percentage of initial species that will see a shift. Panel B zooms in to specific areas shown in grey boxes of A.](../Figures/Fig1.png) -->

## Discussion

>>>>

---
output:
  bookdown::pdf_book:
    toc : no
    keep_tex: no
    number_sections: yes
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
- \floatplacement{table}{H}
- \usepackage{caption}
- \captionsetup[figure]{name = Figure A3}
- \captionsetup[table]{name = Table A3}
---

```{r setup, eval = T, echo=F, warning=F,message=F, results='hide'}
library(MyFunctions)
#### Library ####
packages <- c(
  "dplyr", # Data manipulation
  "knitr",
  "kableExtra",
  "png",
  "grid",
  "janitor"
)
my_lib(packages)
```

```{r FigureS1, eval = T,  echo = F, fig.width=9, fig.height=4, fig.cap =  "World Exclusive Economic Zones used in this study as devided by the Sea Around Us and their centroids (points)."}
